---
title: "Adding parameter heterogeneity to structural models: some guidance for the practitioner^[This project benefited greatly from comments by Glenn Harrison and two anonymous referees.]"
author: "James R. Bland^[Department of Economics, The University of Toledo, 2801 Bancroft St, Toledo, OH 43606 USA, james.bland@utoledo.edu, orcid.org/0000-0002-7117-9998]"
date: "`r Sys.Date()`"
output: 
  bookdown::pdf_document2:
    citation_package: natbib
    global_numbering: true
    keep_tex: true
citation_package: natbib
bibliography: "bib.bib"
biblio-style: "chicago"
toc: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,cache=TRUE)
library(tidyverse)
library(rstan)
library(kableExtra)
library(viridis)
```

> **Abstract**: Participants in economic experiments are heterogeneous, and so it is important to respect this heterogeneity when estimating structural models. I discuss several ways to incorporate between-participant heterogeneity in parameters into structural models. 

# Introduction

When we take a structural model to data from an economic experiment, we are typically interested in estimating the parameters of a model, or using these estimates to construct other derived quantities. One decision the practitioner is faced with when estimating these models is the treatment of heterogeneity in these parameters. It is unlikely, for example, that the parameters of our model are the same for all of our participants. We therefore need to carefully make an appropriate assumption about how we model this heterogeneity. I survey some common approaches to handling heterogeneity in models' parameters. 
Between the two extremes of (i) full pooling of all data and (ii) separately estimating one model for each participant, there is a myriad of options that present their own trade-offs between tractability, computational feasibility, realism, and usefulness. 

I present several possible treatments of heterogeneity in the context of estimating a popular parametric model of expected utility theory (EUT)  from a binary lottery choice experiment [@HS2023].  These models are not intended to be a full treatment of all possible ways one could augment their structural estimation to account for heterogeneity. Rather, they are designed to illustrate some useful building blocks one may wish to use during the construction phase of one's model design.

I use Bayesian techniques to estimate all models discussed below. This is mainly due to the advantages of these techniques over their maximum likelihood counterparts in estimating hierarchical models, especially when using Hamiltonian Monte Carlo [@Betancourt2015]. This certainly does not preclude using maximum likelihood in any of the approaches discussed below. Where Bayesian techniques typically use data augmentation, it is more common for maximum likelihood applications to use *simulated* maximum likelihood.^[See for example @Moffatt2020, chapter  10.] That is, typical data augmentation applications used in Bayesian techniques treat the unobserved participant-level parameters as parameters to be estimated jointly alongside the pooled model of the population. The likelihood function is then conditional on all of the population- and individual-level parameters. On the other hand, one could work with the *unconditional* (of the individual-level parameters) likelihood by integrating them out, usually numerically. This is how simulated maximum likelihood is implemented.^[See @Andersen2012 for an example of this applied to structural models of choice under risk].  

I focus on one important type of heterogeneity: *parameter* heterogeneity. Here, we have *one* model that we would like to estimate, but may want to have some flexibility in how the model's parameters vary between participants. Another important type of heterogeneity is *model* heterogeneity, where we have more than one model that could be generating our data. Of course, within each model, one could augment these with parameter heterogeneity as well.^[See for example the work on mixture models by @HR2009, @Conte2011, and @Gao2025. All of these applications also allow for various kinds of parameter heterogeneity. Another example of model heterogeneity is Strategy Frequency Estimation [@DBF2011], where the typical implementation assumes data are generated by a finite list of strategies (models). Here, parameter heterogeneity is often assumed away, although it certainly can be incorporated [@Bland2020].]

Adding heterogeneity to a structural model, while usually improving its merits in terms of realism, of course comes at the cost of tractability. It is therefore important that we think carefully about the *kind* of heterogeneity we should include in our models. This does not necessarily mean that parameters that we do not care about need not be heterogeneous. For example both @Wilcox2006 and @Bland2021 find that the treatment of a choice precision parameter $\lambda$ in a model, usually a nuisance parameter, can affect descriptive and normative conclusions drawn from that model. 

All of this would be next to useless without example code to replicate the results. As such, please visit https://github.com/JamesBlandEcon/HeterogeneityForThePractitioner for everything needed to reproduce the estimations that appear below.

# Example dataset and model 

I use data from @HS2023 to demonstrate various treatments of participant-level heterogeneity. In particular, I use the subset of these data corresponding to the 175 undergraduates who participated in the "house money" treatment. Each participant made 100 pair-wise choices between a "Left" and "Right" lottery, and at the end of the experiment one decision was selected for payment, and the outcome of this lottery choice was paid to the participant. As the Left and Right lotteries in each pair share the same three prizes, we can describe the experiment and data as follows:

* $x_{i,t}$ is a vector of prizes presented to participant $i$ for their $t$th decision,
* $q^L_{i,t}$ and $q^R_{i,t}$ are vectors of probabilities associated with the Left and Right lotteries presented to participant $i$ for their $t$th decision, and 
* $y_{i,t}$ is an indicator variable equal to one if and only if participant $i$ chose the Left lottery for their $t$th decision, and equal to zero otherwise


I take a popular 2-parameter model of expected utility theory to these data. Specifically I assume (i) a constant relative risk-aversion (CRRA) utility function over certain amounts of money, (ii) logit choice between the lottery pairs, and (iii) the contextual utility normalization [@Wilcox2011]. That is, the model's two fundamental parameters are $r$, the CRRA parameter, and $\lambda$, the logit choice precision parameter.^[Here it is instructive to distinguish between two important but different uses of the word "precision". First, there is the precision *parameter* $\lambda$, which refers to the fundamental parameter of the structural model that describes how deterministic utility differences are translated into the model's probabilistic prediction. Then, there is *statistical* precision in the *estimates*. That is, the inverse of the standard deviation of the posterior distribution (or the standard error if we are in a Frequentist framework) of the parameters in the model. @Andersen2012 discusses this distinction at length in the context of structurally estimating risk preferences. ]
This means that the (normalized) utility *difference* between the Left and Right lotteries as a function of the CRRA parameter $r_i\in \mathbb{R}$ is:

$$
\begin{aligned}
\Delta_{i,t}&=\frac{\sum_{k=1}^Kq_{k,i,t}^L u_i(x_{k,i,t})-\sum_{k=1}^Kq_{k,i,t}^R u_i(x_{k,i,t})}{u_i(\overline x_{i,t})-u_i(\underline x_{i,t})}\\
\text{where: }u_i(x)&=\frac{x^{1-r_i}}{1-r_i},\quad 
\overline x_{i,t}=\max\{x_{i,t}\},\quad 
\underline x_{i,t}=\min\{x_{i,t}\}
\end{aligned}
$$

where the denominator of the first line is the contextual utility normalization, which re-scales utility differences so that they are in the interval $[-1,\ 1]$

The logit choice rule with precision parameter $\lambda_i>0$ then specifies a probability that the participant will choose the Left lottery:


\begin{align}
p(y_{i,t}=1\mid r_i,\lambda_i)&=\Lambda(\lambda_i\Delta_{i,t})
(\#eq:indLikelihood)
\end{align}

where $\Lambda(x)=(1+\exp(-x))^{-1}$ is the inverse logit function. 

While this specification for EUT is frequently found in the literature, it certainly does not represent the cutting edge of models that *could* be taken to these data. @HS2023 in fact take models of  rank-dependent utility [@Quiggin1982], disappointment aversion [@Gul1991], and cumulative prospect theory [@Kahneman2013] to these same data. Instead for pedagogical reasons, I aim to keep the underlying model as simple as possible in order to showcase the various treatments of heterogeneity described below. Furthermore, since this model has only two parameters, all parameter estimates can be easily visualized on a two-dimensional page. 

# Adding heterogeneity to models

## Two extremes: ignoring the heterogeneity and participant-specific estimation

The simplest model we could take to the data is one that assumes that all participants have the same parameters $r_i=r$ and $\lambda_i=\lambda$. This is often referred to as a "pooled" model. In this situation, we can build up the likelihood function from \@ref(eq:indLikelihood) by assuming that each decision is independent conditional on the model's parameters $(r,\lambda)$, that is:

$$
p(y\mid r,\lambda)=\prod_{i=1}^N\prod_{t=1}^T p(y_{i,t}\mid r,\lambda)
$$

Estimation of parameters $r$ and $\lambda$ can then be done by maximum likelihood, or using Bayesian techniques (after combining with a prior). Estimates from the Bayesian approach are shown in Table \@ref(tab:pooledEstimation), which summarizes some properties of the posterior distribution.^[The prior used for this estimation is motivated by @Bland2023, and stated in Appendix A. ] While these estimates are certainly easier to interpret than the models permitting heterogeneity that follow, it should be noted that they may not be representative of *any* kind of measures of central tendency. For example. we can interpret the posterior mean of $r=0.82$, and in fact the entire 2.5%-97.5% range, as representing "very risk averse" risk preferences [@Holt2002]. However this does not mean that we can conclude that all of these participants, or even the *median* participant, has risk preferences at all representative of being "very risk averse". This is in part because the model we have taken to the data is non-linear. But perhaps more importantly, we cannot conclude this because we have estimated a model that has no concept of central tendency in these parameters, as they are assumed to be homogeneous. 



```{r pooledEstimation,echo=FALSE}

TAB<-summary("fits/fit_pooled.rds" |>  readRDS())$summary[c("r","lambda"),
            c("mean","sd","2.5%","25%","50%","75%","97.5%")
]
                                                        
colnames(TAB)<-c("mean","sd","2.5\\%","25\\%","50\\%","75\\%","97.5\\%")

rownames(TAB)<-c("$r$","$\\lambda$")

TAB |>
  kbl(caption = "Summary of the posterior distribution from the pooled model",digits=2,format="latex",booktabs=TRUE,escape=FALSE) |>
  #kable_classic(full_width=FALSE) |> 
  add_header_above(
    c(" "=3,"percentiles"=5)
  )   

```



If we can estimate this pooled model, then in principle we can also estimate a separate model for each participant.  The likelihood for participant $i$'s choices $y_i$ is:


$$
p(y_i\mid r_i,\lambda_i)=\prod_{t=1}^T p(y_{i,t}\mid r_i,\lambda_i)
$$

The posterior mean estimates for all 175 participants in the example dataset are shown in Figure \@ref(fig:individualEstimation). This Figure highlights the extent of heterogeneity that we can expect in economic experiments: our participants are certainly not homogeneous! Furthermore, they are heterogeneous in both the risk-aversion parameter $r$, and the choice precision parameter $\lambda$. Figure \@ref(fig:individualEstimation) also shows the posterior means from the pooled model (i.e. Table \@ref(tab:pooledEstimation)). Here we can see that while the pooled estimate of $r$ (vertical dashed line) is *somewhat* close to the center of the dots (it is in fact slightly to the right of the mean horizontal coordinate), the pooled model *substantially* under-estimates the choice precision parameter $\lambda$ compared to the average of these individual-level estimates. This is typical of many applications, where pooled estimates of choice precision parameters typically imply *less* precise choices than most individual-level estimates. Qualitatively, we can understand this by noting that since the pooled model is forced to fit the same $r$ to all of the data, there is less variation in choices that can be "explained" by $r$. Hence, the model compensates for this by modeling choices as being less precise. 



```{r individualEstimation,echo=FALSE,fig.cap = "Posterior mean estimates from participant-specific models. Posterior means from the pooled model are shown as dashed red lines."}

individual<-"fits/summary_individual.rds" |>
  readRDS() |>
  filter(par!="lp__") |>
  mutate(
    parameter = par,
    parameter = ifelse(parameter=="lambda","\u03bb",parameter)
  ) 

imeans<-individual |> 
  pivot_wider(
    id_cols = id,
    names_from = parameter,
    values_from = mean
  )

(
  ggplot()
  +geom_point(data=imeans,aes(x=r,y=λ))
  +geom_vline(xintercept=TAB[1,1],color="red",linetype="dashed")
  +geom_hline(yintercept=TAB[2,1],color="red",linetype="dashed")
  #+geom_vline(xintercept=mean(imeans$r))
  #+geom_hline(yintercept=mean(imeans$λ))
  #+geom_point(x=TAB["r","mean"],y=TAB["λ","mean"],aes(shape="pooled"),color="red")
  +scale_y_continuous(trans="log10")
  +theme_bw()
  +theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
  +ylab(expression(lambda))
)

```


Practically, estimating one model per participant means that we will be using many fewer observations for every estimation, and so these estimates will be less precise. For example, the example dataset has 17,500 decisions in total, but only 100 decisions per participant. This reduction in precision of the posterior distribution is illustrated in Figure \@ref(fig:pooledSD), which shows the empirical distribution of posterior standard deviations of $r$ for the individual-level estimation. The vertical dashed line at the left of the plot shows the posterior standard deviation from the pooled model, and the vertical dashed line on the right shows the prior standard deviation. 

```{r pooledSD,echo=FALSE, fig.cap="Distribution of posterior standard deviations of individual-level estimates of risk aversion parameter $r$."}

labels<-tibble(
  x = c(0.06,0.34),
  y = c(10,10),
  label = c("pooled model","prior")
)

isds<-individual |> 
  pivot_wider(
    id_cols = id,
    names_from = parameter,
    values_from = sd
  )

#(
#  ggplot()
#  +geom_point(data=isds,aes(x=r,y=λ))
#  +geom_vline(xintercept=TAB["r","sd"],color="red",linetype="dashed")
#  +geom_hline(yintercept=TAB["λ","sd"],color="red",linetype="dashed")
#  +scale_y_continuous(trans="log10")
#  +theme_bw()
#)

(
  ggplot()
  +geom_density(data =individual |> filter(par=="r"),aes(sd))
  +theme_bw()
  +theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
  +geom_text(data=labels,aes(x=x,y=y,label=label))
  +geom_vline(xintercept=c(0.36,0.019),linetype="dashed")
  +xlab("posterior standard deviation")
  +ylab("density")
  +xlim(c(0,0.4))
)

```



While this may be a very *flexible* way to model our data in that each participant has their own parameters, this may not be the most *efficient* way to aggregate what we learn about these parameters from our experiment. That is, when we estimate one model per participant, we are ignoring any information that the choices of participant $j\neq i$ might tell us about participant $i$. 


## Using participant characteristics 

Where participant characteristics are available to us, perhaps from a post-experiment survey, it is fairly straightforward to extend a pooled model to allow its parameters to be functions of these characteristics. @HR2009 for example, estimate models of decision-making under risk whose parameters are a linear function of six characteristics of the participants. The assumption made in this type of specification is that our model's parameters are *deterministic* functions of participant's characteristics. Put differently, if we have two participants with exactly the same characteristics, the model will make exactly the same predictions for these participants. The usefulness of the model, then, depends on how much of the variation in the parameters (in our case $r_i$ and $\lambda_i$) can be "explained" by these characteristics. 

For this application, I use the `age` and `female` variables in the @HS2023 dataset as explanatory variables for both $r_i$ and $\lambda_i$.^[A few participants reported their age as an intyerval. For these participants I use the midpoint of this interval.] That is, the specification for the model's parameters is:

$$
r_i=\beta_{r,0}+\beta_{r,1}\mathrm{female}_i+\beta_{r,2}\mathrm{age}_i
$$

and, using the natural logarithm transform to ensure that the choice precision parameter is positive:

$$
\log\lambda_i=\beta_{\lambda,0}+\beta_{\lambda,1}\mathrm{female}_i+\beta_{\lambda,3}\mathrm{age}_i
$$


```{r characteristics,echo=FALSE}

rounding<-3

characteristics<-summary(
  readRDS("fits/fit_characteristics.rds")
)$summary |>
  data.frame()

characteristics$par<-rownames(characteristics)

characteristics<-characteristics |>
  filter(par!="lp__") |>
  mutate(
    parameter = ifelse(grepl("_r",par),"$r$","$\\log(\\lambda)$"),
    group = c("(constant)","female","age")[parse_number(par)],
    msd = paste0(mean |> round(rounding)," (",sd|> round(rounding),")")
  ) |>
  pivot_wider(
    id_cols = group,
    names_from = parameter,
    values_from = msd
  ) |>
  rename(` `=group)

characteristics |>
  kbl(caption = "Estimates from a model allowing the parameters to vary by participant characteristics. Posterior means with standard deviations in parentheses.",escape=FALSE,booktabs=TRUE) |>
  kable_classic(full_width=FALSE)

```

Importantly here, note that while these equations may look comforting due to their similarity to linear regression equations, there is an important term missing: the error! If most of the heterogeneity in $r_i$ and $\lambda_i$ is left un-explained by these characteristics, then our model will not do much better than the fully pooled model. On the other hand, if there is a relationship between our parameters and these characteristics, then this specification will be able to exploit it. Furthermore, we will be able to infer things about how our model's parameters are related to these characteristics, which may speak directly to our research question. Table \@ref(tab:characteristics) shows posterior estimates from this specification. Assuming that the specification is correct, we can interpret (say) the coefficient on $r$ for the `female` variable as "women are more risk-averse than men of the same age by about 0.2 points". 

We could also use these estimates to put the difference between male and female risk preferences into context. For example, consider the certainty equivalent for a lottery paying \$10 with probability 50\%, and \$50 otherwise. 
This certainty equivalent is:
$$
\begin{aligned}
C&=\left(0.5\times 10^{1-r}+0.5\times 50^{1-r}\right)^{\frac{1}{1-r}}
\end{aligned}
$$

A 20 year-old male participant would have a certainty equivalent of  about \$24.50, whereas a 20 year-old female participant would have a certainty equivalent of about \$24.49.^[This is just using the posterior mean estimates from Table \@ref(tab:characteristics). A full posterior distribution over these certainty equivalents is of course calculable from the posterior simulation of the model's fundamental parameters. ]






## Finite mixtures 

A Pearson-style finite mixture specification assumes that the data as a whole are generated by $C$ discrete components, which I will index as $c\in\{1, 2, \ldots, C\}$. Each of these components specifies a data-generating process conditional on an observation being drawn from that component, which we can notate in our example as $p(y_{i,t}\mid r_c,\lambda_c)$, which is the probability (likelihood) of observing action $y_{i,t}$ given that it was drawn from component $c$ of the distribution, whose parameters $\{r_c,\lambda_c\}$ describe the behavior. *Mixing probabilities* $\{\rho_c\}_{c=1}^C$ then describe  the fraction of the whole distribution that is generated from each component. Examples of applications of finite mixture models include @Bruhin2010 and @BFS2019.


Here, it is important to distinguish between mixtures of *parameters* and mixtures of *models*. This section is specifically discussion the former, where the underlying model generating the data is the same for all participants, but the parameters that go into that model are different between participants. While not discussed further here, it is important to acknowledge the use of the word "mixture" in the literature to mean (at least) two other things, which relate to how different *models* are combined in these kinds of mixture models. As documented in Section 1 of @Gao2025, there is the random preference mixture of *choices*, in which decision-makers draw a fresh preference *functional* for each task. That is, participants are different because the functional form of their utility is different, not (just) that the parameters in the function differ between participants.  This is the kind of mixture model studied in, for example, @Loomes1995, @Loomes1998, and @HR2009.^[A special case of this kind of model, as implemented in @Conte2011, is one where participants have just one draw of their preference functional, and use it for the entire experiment.] There is also the "genetic" mixture model or "mule" specification [@Harrison2026], where the agents' utility function itself is a convex combination of two or more component functionals. Of course, any of these other kinds of mixture model could have their own treatment of *parameter* heterogeneity. In fact many of the above-cited examples do so: @Gao2025 also has a Bayesian hierarchical structure (discussed in Section 3.4), @HR2009 uses detrministic linear functions of observable participant characteristics (discussed in Section 3.2), and @Conte2011 has a hierarchical flavor (in that participants' parameters are modeled as draws from a population distribution), although it is implemented using simulated maximum likelihood estimation. 
These kinds of mixture model deserve their own treatments, but the remainder of this section will be devoted solely to the Pearson-style mixture model, and its use in accounting for *parameter* heterogeneity. 

Given that economic experiments typically have their participants make more than one decision, once one decides on a mixture specification, one must also decide *at what level* this mixing occurs. For most applications, this usually comes down to deciding whether the mixing happens at the *participant* level, or the *decision* level. This choice affects how the likelihood function is written down, and perhaps more importantly, it changes our interpretation of the model. If the mixing happens at the decision level, then decisions made by the same participant are permitted to come from any of the component distributions. In this case if we assumed some  parameters for our model, then we could simulate data from our experiment by repeating the following procedure:

1. Draw a component $c\sim\mathrm{Categorical}(\rho)$
2. Draw data $y_{i,t}$ conditional on the parameters for this component ($r_c$ and $\lambda_c$).

and we could construct the "grand" log-likelihood of the entire model as follows:

$$
\log p(y\mid \rho,\{r_c,\lambda_c\}_{c=1}^C)=\sum_{i=1}^N\sum_{t=1}^T\log\left(\sum_{c=1}^C\rho_kp(y_{i,t}\mid r_c,\lambda_c)\right)
$$


If the mixing happens at the participant level, then the assumption is that data for each participant can only be generated from one of the components. In this case we could simulate data for our experiment as follows:

1. Draw a component $c\sim\mathrm{Categorical}(\rho)$
2. Draw all data for participant $i$ $\{y_{i,t}\}_{t=1}^T$ conditional on this component, from the distribution $y_i\mid r_c,\lambda_c$.

and we can construct the grand log-likelihood of the entire model as follows:

$$
\log p(y\mid \rho,\{r_c,\lambda_c\}_{c=1}^C)=\sum_{i=1}^N\log\left(\sum_{c=1}^C\rho_c\prod_{t=1}^T p(y_{i,t}\mid r_c,\lambda_c)\right)
$$

The finite mixture model lends itself to accounting for heterogeneity by grouping participants (if we are mixing at the participant level) into $C$ categories, each represented by a different set of parameters. For our example of @HS2023, we can achieve this by assuming that there are $C$ groups of participants, where each group's behavior is described by parameters $(r_c,\lambda_c)$. Posterior estimates for the 2- and 3-component models are summarized in Figure \@ref(fig:plotMixture). 

```{r tabMixture,echo=FALSE,warning=FALSE}

rounding<-2

TAB<-list()

plotThis<-tibble()


files<-list.files(path="fits")
files<-files[grepl("_mix_",files)]
 


for (ff in files) {
  
  mixSum<-summary(readRDS(paste0("fits/",ff)))$summary |>
    data.frame()
   
  mixSum$par<-rownames(mixSum) 
  
  mixSum<-mixSum |>
    mutate(
      parameter = gsub("[^a-zA-Z]", "", par),
      component = parse_number(par),
      msd = paste0(mean |> round(rounding)," (",sd |> round(rounding),")")
    ) |>
    filter(par != "lp__") |>
    mutate(parameter = 
             ifelse(parameter=="lambda","\u03bb",parameter),
           parameter =  
             ifelse(parameter=="mix","\u03c1",parameter),
           numComponents = max(component)
             ) 
  plotThis<-rbind(plotThis,mixSum)
  
    TAB[[ff]]<-mixSum |> 
      pivot_wider(
      id_cols = component,
      names_from = parameter,
      values_from = msd
    ) |>
    select(-component)
  
}
if (FALSE) {  
  TAB |> 
    kbl(caption = "posterior means (standard deviations in parentheses) of the finite mixture models") |>
    kable_classic(full_width=FALSE) |>
    add_header_above(
      paste(1:length(TAB)+1,"components")
    )
}

```

```{r plotMixture,fig.cap = "Posterior mean estimates from the finite mixture models. Posterior means from the particiapant-specific estimation are shown in black.  ",echo=FALSE}

plotThis<-plotThis |>
  pivot_wider(
    id_cols = c(numComponents,component),
    names_from = parameter,
    values_from = mean
  ) 

( 
  ggplot(data = plotThis,
         aes(x=r,y=`λ`,size=`ρ`,color=as.factor(numComponents))
         )
  +geom_point(data=imeans,aes(x=r,y=`λ`),size=0.5,color="black",alpha=0.2)
  +geom_point()
  +scale_color_manual(values = c("red","blue","green","magenta"))
  +theme_bw()
  +theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
  +scale_size(name="Mixing probability")
  +scale_y_continuous(trans="log10")
  +labs(color = "number of components")
  +ylab(expression(lambda))
)

```



While these estimates may be of interest to the reader on their own, for those wishing to apply these techniques to their own work, the process of how these estimates were calculated should be of more pedagogical use. In particular, note that this kind of mixture model suffers from an *label-switching* problem. For example, for the two-component mixture model presented above, the likelihood function has the following property:

$$
p(x\mid r_1,\lambda_1,\rho_1,r_2,\lambda_2,\rho_2)=p(x\mid r_2,\lambda_2,\rho_2,r_1,\lambda_1,\rho_1)
$$

In fact, for a model with $C$ components, the likelihood function will be maximized at $C!$ different points. For Bayesian inference, this presents a problem because posteriors *must* be multi-modal. Likewise for maximum likelihood estimation, the likelihood function has multiple global maxima. For the Bayesian implementation shown above, this problem was circumvented by imposing that parameter $r$ was ordered from most negative (component 1) to most positive (component $C$). 

Finally, and perhaps most importantly, it is important to question whether our data are actually well-organized by *any* finite mixture. Inspection of the black dots in Figure \@ref(fig:plotMixture) shows (to this author, at least) a much more *continuous* distribution of types than any small number of components could reasonably approximate. Interpreting the estimated types as being "economically relevant" is therefore not recommended, as it could lead the researcher into concluding that the data were "well-organized" by just a few discrete types. 



## Hierarchical models

Hierarchical models perhaps provide a happy medium between full pooling and participant-specific estimation. In a hierarchical model, participants' parameters are heterogeneous, and treated as unknown quantities to be estimated jointly alongside "hyper-parameters" governing the their distribution in the population. An *exchangability* assumption is made that states that prior to observing the data, beliefs about these data do not depend on the units (participants) being observed. Hierarchical models can be useful for (at least) two purposes. Firstly, in estimating the parameters of the population distribution, we are able to make statements about the distribution of parameters in our population. Secondly, as we are also estimating each participant's parameters, these individual-level estimates are informed by the population distribution. As such, we are able to make inference about a specific individual using the information provided by the entire experiment. 

Examples of hierarchical models applied to structural modelling of economic experiments include @Nilsson2011, @Gao2023, and  @Bland2024. Hierarchical models, and in particular *Bayesian* hierarchical models are becoming much more common in the literature, perhaps due to improved computer hardware, and the software that implements it (e.g. @Stan). 

Perhaps the simplest hierarchical model one could estimate with the @HS2023 data is one that assumes parameters $r_i$ and $\lambda_i$ are drawn from a multivariate normal distribution as follows:

$$
\begin{pmatrix}r_i & \log \lambda_i\end{pmatrix}^\top \sim iidN(\mu,\Sigma)
$$

where we take the natural logarithm of $\lambda_i$ to ensure that the parameter is strictly positive. Here $\mu$ is a 2-vector of population means, and $\Sigma$ is a $2\times 2$ variance-covariance matrix. Note that $\Sigma$ can be de-composed into a vector $\tau$ of standard deviations and a correlation matrix $\Omega$ such that:^[This permits assigning hyper-priors to the standard deviations and correlations separately. See the *Stan User's Guide* for details here: https://mc-stan.org/docs/stan-users-guide/regression.html#multivariate-hierarchical-priors.section.]

$$
\Sigma=\mathrm{diag\_matrix}(\tau)\Omega \mathrm{diag\_matrix}(\tau)
$$

Table \@ref(tab:hierarchicalTable) summarizes the posterior distribution of these population-level parameters from the hierarchical model, and we can see the individual-level parameter estimates (posterior means) in Figure \@ref(fig:hierarchicalIndividual). Here we should note that since we are doing Bayesian estimation, these are simply summary statistics of the full posterior simulation of the hierarchical model's parameters. This underlying posterior simulation is of course useful and easy to access for every parameter in the hierarchical model, whether we are interested in a population-level parameter (like $\mu$) or a subject-level parameter like $r_i$. For example Figure \@ref(fig:subject1) shows all 20,000 posterior draws (post warm-up) of $(r_i,\lambda_i)$ for the first two subjects that appear in the data. Here we can see that Subject 1 (red) is both more risk-averse and makes more precise choices compared to Subject 2 (blue). We can see this in the plot as the red dots mostly being to the right of and above the blue dots. We can use these draws to compute many things. For example, we can compute the posterior probability (up to simulation accuracy) that Subject 1 is more risk-averse than Subject 2 by simply calculating the fraction of red dots that are to the right of their corresponding blue dots, which is $99.9\%$.^[Here we need to make sure that we only compare draws for each subject that come from the same simulation iteration.] It should be noted here that it is not uncommon for the marginal posterior distributions of some of these subject-level parameters to be skewed or bi-modal. This is not a bug of the Bayesian Hierarchical model, but rather the model telling us that there might be substantial uncertainty associated with these parameters. Alternatively for multiple modes, it may be that the combination of experiment design and participant's decisions could be "explained" by the model roughly equally well by multiple combinations of parameter values. 




    
    
```{r hierarchicalTable,echo=FALSE,warning=FALSE,eval=TRUE} 
hierarchical<-summary(readRDS("fits/fit_hierarchical.rds"))$summary |>
  data.frame()

hierarchical$par<-rownames(hierarchical)

hierarchical<-hierarchical |>
  mutate(
    parameter = gsub("[^a-zA-Z]", "", par),
    component = parse_number(par)
  ) |>
  filter(parameter!="z")

hierarchical_shrinkage<-hierarchical |>
  filter(parameter=="r" | parameter == "lambda")

rounding<-3

hierarchical_population<-hierarchical |>
  filter(
    parameter=="MU" | parameter == "TAU" | par=="OMEGA[1,2]"
  ) |>
  mutate(
    msd = paste0(mean |> round(rounding)," (",sd |> round(rounding),")"),
    indparam = c("$r$","$\\log(\\lambda)$")[component],
    popparam = "$\\mu$ (mean)",
    popparam = ifelse(parameter=="TAU","$\\tau$ (sd)",popparam),
    popparam = ifelse(parameter=="OMEGA","correlation",popparam)
  ) 
  
TAB<-hierarchical_population |>
  filter(parameter!="OMEGA") |>
  pivot_wider(
    id_cols = popparam,
    names_from = indparam,
    values_from = msd
  )

colnames(TAB)<-c("",colnames(TAB)[2:3])

TAB<-rbind(TAB,c("correlation",hierarchical_population["OMEGA[1,2]","msd"],""))

TAB |>
  kbl(caption = "Population-level estimates from the hierarchical model. Posterior means with standard deviations in parentheses.",booktabs=TRUE,escape=FALSE) 

```


```{r hierarchicalIndividual,fig.cap="Shrinkage estimates (posterior means) from the hierarchical model.",echo=FALSE,eval=TRUE}

d<-hierarchical_shrinkage |>
  pivot_wider(
    id_cols = component,
    names_from = parameter,
    values_from = mean
  )

(
  ggplot(d,aes(x=r,y=lambda))
  +geom_point()
  
  +scale_y_continuous(trans="log10")
  +ylab(expression(lambda))
  +theme_bw()
  +theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
)

```

```{r subject1, fig.cap = "20,000 posterior draws from the posterior distribution of the parameters for Subjects 1 and 2 from the hierarchical model.",echo=FALSE,eval=TRUE}

Fit<-"fits/fit_hierarchical.rds" |>
  readRDS()

R<-extract(Fit)$r
LAMBDA<-extract(Fit)$lambda

rlambda<-data.frame()


for (ii in c(1, 2)) {
  rlambda<-rbind(
    rlambda,
    data.frame(
    r = R[,ii],
    lambda = LAMBDA[,ii],
    subj = paste("Subject",ii)
  ) |>
    mutate(
      draw = 1:n()
    )
  )
}

(
  ggplot(rlambda, aes(x=r, y=lambda,color=subj))
  +geom_point(size=0.2,alpha=0.1)
  +guides(color = guide_legend(override.aes = list(size = 5, alpha=1)))
  +scale_color_manual(name="", values = c("red","blue"))
  #+facet_wrap(~subj)
  +xlab("r")
  +ylab(expression(lambda))
  +theme_bw()
  +theme(axis.title.y = element_text(angle = 0, vjust = 0.5))
  +theme(legend.position="top")
  +scale_y_continuous(trans="log10")
)

# for the post prob that Subj 1 is more risk averse

d<-rlambda |>
  pivot_wider(
    id_cols  = draw,
    names_from = subj,
    values_from = r
  ) |>
  ungroup() |>
  summarize(
    mean(`Subject 1`>`Subject 2`)
  )


```

Even if one is not directly interested in participant-specific heterogeneity, one may still want to estimate a hierarchical model. For example one may be interested just in the population means parameter $\mu$. However as structural models are often nonlinear, they can suffer from *heterogeneity bias*: if we assume a pooled model when parameters are in fact heterogeneous, then we will likely have a biased estimator. For example  @Wilcox2006 and @Bland2024 stress the importance of modeling heterogeneity, even when the inferential goal is about average behavior. On the flip side, even if we are not interested in the population-level parameters ($\mu$ and $\Sigma$), hierarchical models are useful if we want to make inference about individuals. Here, we leverage the entire dataset to form informative priors (represented by $\mu$ and $\Sigma$) for our participants' parameters.

# Conclusion

Our participants are heterogeneous, and so we should allow for heterogeneity in our structural models. The previous section outlines various approaches to doing this. Whether we are interested in making statements about our population, or interested in making good inference about an individual, we have a wide variety of tools to allow for both discrete and continuous types of heterogeneity in our models. The appropriateness of each option are, of course, application-specific. For example, if one is interested in how parameters vary with participants' characteristics, or suspects that there to be a strong correlation between characteristics and parameters, then allowing parameters to be functions of characteristics makes sense. On the other hand, if this kind of pooling assumption seems too restrictive, then perhaps a hierarchical model is more appropriate. 

The above discussion, of course, does not provide a complete treatment of heterogeneity. Rather, it outlines some popular starting points the practitioner could consider, and the various merits of each one. Neither are any of the methods discussed here mutually exclusive. For example it is entirely feasible to estimate a hierarchical model where the population-level parameters vary by participant characteristics. Alternatively, a hierarchical model could assume a finite mixture of multivariate normal distributions is generating the participant-specific parameters. 

# Appendices {-}

## A -- Priors {-}

### Pooled and individual estimation {-}

In the pooled and individual estimation, I assume the following priors calibrated in @Bland2023:

$$
\begin{aligned}
r&\sim N(0.27,0.36^2)\\
\log \lambda &\sim N(\log(30),0.5^2)
\end{aligned}
$$

### Participant characteristics {-}

For the pooled estimation with participant characteristics, the constant terms inherit the priors for $r$ and $\lambda$ from the pooled estimation. The priors for coefficients for the characteristics have a prior distribution of $N(0,0.1^2)$.

### Finite mixtures {-}

Each component of the mixture has priors the same as for the pooled estimation, except that the vector for $r$ is restricted to being strictly increasing (i.e. $r_1<r_2<\ldots <r_C$). The mixing probability prior is:

$$
\rho \sim \mathrm{Dirichlet}(\boldsymbol 1)
$$

where $\boldsymbol 1$ is a vector of ones the same size as the number of components. 


### Hierarchical {-}

For the hierarchical model, the priors for the pooled estimation become the priors for the population mean parameter $\mu$, i.e.:

$$
\begin{aligned}
\mu_r&\sim N(0.27,0.36^2)\\
\mu_\lambda &\sim N(\log(30),0.5^2)
\end{aligned}
$$


For the standard deviation parameters I assume $\tau\sim \mathrm{Cauchy}^+(0,1)$, and for the correlation matrix $\Omega\sim LKJ(4)$.




# References {-}



